Static Race Detection in OS Kernels by Mining Locking Rules

### 2023-08-02 5.evaluation.tex
Finish a temporary version (without comparing experiment and some evaluation results)

\begin{table}[tbph]
	\tablecaption{Detection results of Linux 6.2.}
	\label{tbl_bug_detection}
	\renewcommand{\arraystretch}{1}
	\setlength\tabcolsep{2pt}
	\noindent{\scriptsize
		\begin{center}
			\begin{tabular}{p{1.5cm}|l|c}
				\hline
				\multicolumn{2}{c|}{\textbf{Description}} & \textbf{\sys}  
				\\ \hline
				\multirow{2}{1.5cm}{\textbf{{\em Code analysis}}} & 
				Source files (analyzed/all) & xxxK/xxxK
				\\ \cline{2-3}
				& Source code lines (analyzed/all) & xxxM/xxxM 
				\\ \cline{2-3}
				\hline
				\multirow{3}{1.5cm}{\textbf{{\em Locking-rule mining}}} & 
				Key fields / total variables& 
				\\ \cline{2-3}
				& Mined locking rules &
				\\ \cline{2-3}
				& Protected field accesses / all accesses & 0.6
				\\ \cline{2-3}
				\hline
				\multirow{2}{1.5cm}{\textbf{{\em Data race detection}}} & 
				Detected data races (real / all) & 273 / 341
				\\ \cline{2-3}
				& Dropped data races by lock-usage analysis & 63
				\\ \cline{2-3}
				\hline
				\multirow{4}{1.5cm}{\textbf{{\em Data race estimation}}}
				& Null-pointer dereference (confirmed / all) & 15 / 20
				\\ \cline{2-3}
				%& Infinite loop (confirmed / all) & 0 / 1
				%\\ \cline{2-3}
				& Data inconsistency (confirmed / all) & 7 / 10
				\\ \cline{2-3}
				& Unprotected write (confirmed / all) & 10 / 57
				\\ \cline{2-3}
				& Total harmful data races (confirmed / all) & 32 / 88
				\\ \cline{2-3}
				\hline
				\multirow{3}{1.5cm}{\textbf{{\em Time usage}}} & 
				Key-field extraction & 
				\\ \cline{2-3}
				& Data-race detection &
				\\ \cline{2-3}
				& Total time &
				\\ \cline{2-3}
				\hline
			\end{tabular}
	\end{center}}
\end{table}

@@

We configure \sys with common lock-acquiring/release functions (like {\tt 
spin\_lock and spin\_unlock}) to perform lock-set analysis to extract key 
fields and mine locking rules, and lock-initialization functions (like {\tt 
spin\_lock\_init}) to filter our false data races caused by code paths that 
can not execute concurrently. And then run \sys to automatically check the 
kernel source code. We manually check all the data races found by \sys, and 
Table~\ref{tbl_bug_detection} shows the results, and source code lines are 
counted by CLOC~\cite{cloc}. From the results, we have the following findings:

@@

\PP{Locking-rule mining.} An OS kernel has a large code base with numerous 
variables. Handling all variables when mining locking rules can introduce much 
overhead. However, we observe that a variable tends to be protected by the lock 
stored in the same data structure as the accessed variable. Based on this 
observation, our locking-rule mining method first extracts a key field by 
finding whether there exists any access to it that is protected by a lock 
stored in the same data structure. This method drops xxx\% variables (xxx out 
xxx) that need to be handled when mining locking rules, and thus can reduce 
overhead significantly. After extracting key fields, our locking-rule mining 
method collects all accesses to these key fields, and then deduces locking-rule 
based on statistics. In this paper, given a data structure field, we set the 
threshold of the ratio of accesses protected by a specific lock to all accesses 
to 0.6. Our alias-aware rule mining method can drop many false rules and thus 
only mines xxx rules, which can effectively reduce false data races.

\PP{Data race detection.} \sys reports 341 data races in the kernel source 
code. We spent 15 hours on checking these data races and identify that 273 of 
them are real, with a false positive rate of 19.9\%. Besides, our lock-usage 
analysis drops 63 false data races and the results show that it can reduce 
false positives significantly.

\PP{Data race estimation.} Many data races are benign and can not cause memory 
or logic bugs, and thus developers are unwilling to put effort into repairing 
them. We exploit three patterns to detect null-pointer dereferences, data 
inconsistency and unprotected write as introduced in 
Section~\ref{subsec_estimation}, and find 88 data races in these patterns. We 
report them to developers and 32 of them have been confirmed and fixed by them. 
We still wait for the response of other data races. Moreover, one of the 
developers wonders if \sys can be used in their CI to detect these problems. 
The results show that our pattern-based estimation can extract harmful data 
races effectively, and can considerably reduce the workload of developers.

@@

Second, in order to reduce memory overhead and improve the performance of data 
passing among different functions, an integer can be divided into several bit 
vectors to represent different data structure fields. However, in the LLVM 
bytecode, accesses to these vectors are divided into a LOAD operation and 
several bit operations, and thus \sys can not distinguish accesses to different 
fields and reports false data races, because not all these fields should be 
protected by a lock. This reason causes \sys to report 14 false data races.

Apart from false data races caused by the two main reasons, there are still 7 
other data races. \sys regards a function without caller functions in a kernel 
module as the entry function and starts analysis from these entries. However, 
in 4 cases, an assertion is put at the beginning of the entry function, to 
guarantee that the required lock is held, but \sys does not consider assertions 
and the subsequent accesses are regarded as unprotected. The other 3 false data 
races are accessed with {\em READ\_ONCE} or {\em WRITE\_ONCE}, and can not 
cause serious issues.

@@

Finally, \sys exploits a lock-usage analysis to filter out false data races 
caused by code paths that can not execute concurrently. Specifically, the 
lock-usage analysis extracts all functions that are reachable from the lock 
initialization functions in a call graph with the union-find set, and assumes 
that these functions can not execute concurrently. However, some functions can 
be called at multiple program sites, and some of the program sites may be not 
reachable from the lock initialization functions, but data races that exist in 
these functions are all dropped by our lock-usage analysis.

@@

\PP{Null-pointer dereference in the HDA sound driver.} In 
Figure~\ref{fig_case_bugs}(a), the functions {\em snd\_hdac\_regmap\_exit()} 
and {\em snd\_hdac\_regmap\_sync()} can execute concurrently. In Thread 2, the 
variable {\em codec->regmap} is checked by an if statement in the function {\em 
snd\_hdac\_regmap\_sync()}. Right after the condition of the if statement is 
calculated to be true, the variable {\em codec->regmap} is set to be NULL by 
the function {\em snd\_hdac\_regmap\_exit()} in Thread 2. And then the function 
{\em regcache\_sync()} is called in Thread 1, with the argument {\em 
codec->regmap}, after acquiring the lock {\em codec->regmap\_lock}. In the 
called function, the variable {\em codec->regmap\_lock} is dereferenced through 
{\em map->lock()}. In this execution case, the data structure field {\em 
codec->regmap} is first assigned with NULL in Thread 1, and then dereferenced 
in Thread 2, and thus cause a null-pointer dereference.

@@

\PP{Unprotected write in the LPFC SCSI driver.} In 
Figure~\ref{fig_case_bugs}(c), the value of the shared variable {\em 
phba->fcf.fcf\_flag} is set to 0 by the function {\em 
lpfc\_unregister\_fcf\_rescan()}, without acquiring the lock {\em 
phba->hbalock}, and this can cause double-fetch in several program sites. For 
example, if {\em phba->fcf.fcf\_flag} is set to 0 by Thread 2 between the two 
accesses to it in the function {\em lpfc\_register\_fcf()} in Thread 1 (Lines 
1873 and 1874), the values of the two accesses can be different, causing 
unpredictable behaviors.

>>>>

\begin{table}[tbph]
	\tablecaption{Detection results of Linux 6.2.}
	\label{tbl_bug_detection}
	\renewcommand{\arraystretch}{1}
	\setlength\tabcolsep{2pt}
	\noindent{\scriptsize
		\begin{center}
			\begin{tabular}{p{1.5cm}|l|c}
				\hline
				\multicolumn{2}{c|}{\textbf{Description}} & \textbf{\sys}  
				\\ \hline
				\multirow{2}{1.5cm}{\textbf{{\em Code analysis}}} & 
				Source files (analyzed/all) & xxxK/xxxK
				\\ \cline{2-3}
				& Source code lines (analyzed/all) & xxxM/xxxM 
				\\ \cline{2-3}
				\hline
				\multirow{2}{1.5cm}{\textbf{{\em Locking-rule mining}}} & 
				Key fields / total variables& 
				\\ \cline{2-3}
				& Mined locking rules &
				\\ \cline{2-3}
				\hline
				\multirow{2}{1.5cm}{\textbf{{\em Data race detection}}} & 
				Detected data races (real / all) & 273 / 341
				\\ \cline{2-3}
				& Dropped data races by lock-usage analysis & 63
				\\ \cline{2-3}
				\hline
				\multirow{4}{1.5cm}{\textbf{{\em Harmfulness estimation}}}
				& Null-pointer dereference (confirmed / all) & 15 / 20
				\\ \cline{2-3}
				& Data inconsistency (confirmed / all) & 7 / 10
				\\ \cline{2-3}
				& Double fetch (confirmed / all) & 10 / 57
				\\ \cline{2-3}
				& Total harmful data races (confirmed / all) & 32 / 88
				\\ \cline{2-3}
				\hline
				\multirow{3}{1.5cm}{\textbf{{\em Time usage}}} & 
				Key-field extraction & 
				\\ \cline{2-3}
				& Data-race detection &
				\\ \cline{2-3}
				& Total time &
				\\ \cline{2-3}
				\hline
			\end{tabular}
	\end{center}}
\end{table}

@@

We configure \sys with common lock-acquiring/release functions (like {\tt 
spin\_lock} and {\tt spin\_unlock}) to perform lock-set analysis to extract key 
fields and collect field accesses, and lock-initialization functions (like {\tt 
spin\_lock\_init}) to filter our false data races caused by code that can not 
execute concurrently. And then run \sys to automatically check the kernel 
source code. We manually check all the data races found by \sys, and 
Table~\ref{tbl_bug_detection} shows the results, and source code lines are 
counted by CLOC~\cite{cloc}. From the results, we have the following findings:

@@

\PP{Locking-rule mining.} An OS kernel has a large code base with numerous 
variables. Handling all variables when mining locking rules can introduce much 
overhead. However, we observe that the variable that is accessed and the lock 
to protect the access often exist in the same data structure. Based on this 
observation, our locking-rule mining method extracts key fields by finding 
whether there exists any access to it that is protected by a lock stored in the 
same data structure. This method drops xxx\% variables (xxx out xxx) that need 
to be handled when mining locking rules, and thus can reduce overhead 
significantly. After extracting key fields, our locking-rule mining method 
collects all accesses to these key fields, and then deduces locking rules based 
on statistics. In this paper, given a data structure field, we set the 
threshold of the ratio of accesses protected by a specific lock to all accesses 
to 0.6. Our alias-aware rule mining method can drop many false rules and thus 
only mines xxx locking rules, which can effectively reduce false data races.

\PP{Data race detection.} \sys reports 341 data races in the kernel source 
code. We spent 15 hours on checking these data races and identify that 273 of 
them are real, with a false positive rate of 19.9\%. Besides, our lock-usage 
analysis drops 63 false data races and the results show that this strategy can 
reduce false positives significantly.

\PP{Data race estimation.} Many data races are benign and can not cause memory 
or logic bugs, and thus developers are unwilling to put effort into repairing 
them. We exploit three patterns to detect null-pointer dereferences, data 
inconsistencies and double fetches as introduced in 
Section~\ref{subsec_estimation}, and find 88 data races in these patterns. We 
report them to developers and 32 of them have been confirmed and fixed by them. 
We still wait for the response of other data races. Moreover, one of the 
developers wonders if \sys can be used in their CI to detect these problems. 
The results show that our pattern-based estimation can extract harmful data 
races effectively, and can considerably reduce the workload of developers.

@@

Second, in order to reduce memory overhead and improve the performance of data 
passing among different functions, an integer can be divided into several bit 
vectors to represent different data structure fields. However, in the LLVM 
bytecode, accesses to these vectors are divided into a load operation and 
several bit operations, and thus \sys can not distinguish accesses to different 
fields and reports false data races, because not all these fields should be 
protected by a lock. This reason causes \sys to report 14 false data races.

Apart from false data races caused by the two main reasons, there are still 7 
other data races. \sys regards functions without caller functions in a kernel 
module as the entry functions and starts analysis from these entries. However, 
in 4 cases, an assertion is put at the beginning of the entry function, to 
guarantee that the required lock is held, but \sys does not consider assertions 
and the subsequent accesses are regarded as unprotected. The other 3 false data 
races are accessed with {\em READ\_ONCE} or {\em WRITE\_ONCE}, and can not 
cause serious issues.

@@

Finally, \sys exploits a lock-usage analysis to filter out false data races 
caused by code that can not execute concurrently. Specifically, the lock-usage 
analysis extracts all functions that are reachable from the lock initialization 
functions in a call graph with the union-find set, and assumes that these 
functions can not execute concurrently. However, some functions can be called 
at multiple program sites, and some of the program sites may be not reachable 
from the lock initialization functions, but data races that exist in these 
functions are all dropped by our lock-usage analysis.

@@

\PP{Null-pointer dereference in the HDA sound driver.} In 
Figure~\ref{fig_case_bugs}(a), the functions {\em snd\_hdac\_regmap\_exit()} 
and {\em snd\_hdac\_regmap\_sync()} can execute concurrently. In Thread 2, the 
variable {\em codec->regmap} is checked by an if statement in the function {\em 
snd\_hdac\_regmap\_sync()}. Right after the condition of the if statement is 
calculated to be true, the variable {\em codec->regmap} is set to be NULL by 
the function {\em snd\_hdac\_regmap\_exit()} in Thread 1. And then the function 
{\em regcache\_sync()} is called in Thread 2, with the argument {\em 
codec->regmap}, after acquiring the lock {\em codec->regmap\_lock}. In the 
called function, the variable {\em codec->regmap} is dereferenced through 
{\em map->lock()}. In this execution case, the data structure field {\em 
codec->regmap} is first assigned with NULL in Thread 1, and then dereferenced 
in Thread 2, and thus cause a null-pointer dereference.

@@

\PP{Double fetch in the LPFC SCSI driver.} In Figure~\ref{fig_case_bugs}(c), 
the value of the shared variable {\em phba->fcf.fcf\_flag} is set to 0 by the 
function {\em lpfc\_unregister\_fcf\_rescan()} in Thread 2, without acquiring 
the lock {\em phba->hbalock}, and this can cause double fetches in several 
program sites. For example, if {\em phba->fcf.fcf\_flag} is set to 0 by Thread 
2 between the two accesses to it in the function {\em lpfc\_register\_fcf()} in 
Thread 1 (Lines 1873 and 1874), the values of the two accesses can be 
different, causing unpredictable behaviors.

### 2023-08-02 6.discussion.tex
Finish a temporary version (without comparing experiment and some evaluation results)

\PP{Benefits in other aspects.} \sys uses an alias-aware mining method to 
deduce precise locking rules about whether accesses to a variable should be 
protected by a lock, and which lock is required. Such rules can benefit in 
various aspects.

First, with these rules, developers can improve the reliability 
and performance of kernel code. On the one hand, developers can carefully check 
whether necessary locks are used when writing code, and avoid serious bugs 
caused by data races in the early developing stage. On the other hand, 
developers can limit the critical zone protected by a lock/unlock pair to the 
exact accesses that should be protected, and let other code can execute 
concurrently. And these locking rules can also help improve documents of OS 
kernels, which is beneficial  for long term evolution of kernel codes.

@@

\PP{Comparison to other approaches.} We have tried our best to experimentally 
compare \sys to existing works in data race detection including 
RELAY~\cite{Voung:FSE07}, WHOOP~\cite{Deligiannis:ASE15} and 
Goblint~\cite{Vojdani:ASE16}. Unfortunately, they are too old and generate too 
many errors when compiling and running, and thus we compare \sys to them 
methodologically.

\PP{Limitations and future works.} \sys can be improved in some aspects. First, 
our alias-aware mining method only considers accessed variables and protecting 
locks that exist in the same data structure, and can miss some locking rules 
involving special locks such as RCU locks. To overcome this limitation, we plan 
to create a common virtual node for RCU lock in the alias graph, and regard it 
as the ancestor of all other nodes. In this way, all variables have a common 
ancestor with the RCU lock, and can be inferred to be protected by the RCU 
lock, because the accessed variable and the RCU lock are existing in the 
identical data structure by our alias-aware rule mining method. Second, \sys 
does not handle function-pointer calls, and thus it can not build complete call 
graphs, and thus can miss data races reached from function-pointer calls. To 
relieve this limitation, we plan to apply existing function-pointer 
analysis~\cite{Hind:TOPLAS99, Zhang:ISPEC15, Heintze:PLDI01} in \sys, to detect 
more data races in functions that are called through function pointers and 
reduce false negatives. Third, \sys performs a path-based analysis and suffers 
from high time overhead, to accelerate its analysis, we plan to employ 
summer-based analysis~\cite{Bai:ATC19, Bai:ATC22} to avoid re-analysis of the 
same function definition. Finally, we plan to apply \sys to detect data races 
in other OS kernels such as FreeBSD~\cite{FreeBSD} and NetBSD~\cite{NetBSD}, 
and use it to detect more types of concurrency bugs including use-after-free 
and uninitialized-variable access.

>>>>

\PP{Benefits in other aspects.} \sys uses an alias-aware mining method to 
deduce precise locking rules about whether accesses to a specific data 
structure field should be protected and which data structure field the 
protecting lock exist in. Such rules can benefit in various aspects.

First, with these rules, developers can improve the reliability 
and performance of kernel code. On the one hand, developers can carefully check 
whether necessary locks are used when writing code, and avoid serious bugs 
caused by data races in the early developing stage. On the other hand, 
developers can limit the critical zone protected by a lock/unlock pair to the 
exact accesses that should be protected, and let other code execute 
concurrently. And these locking rules can also help improve documents of OS 
kernels, which is beneficial  for long term evolution of kernel codes.

@@

%\PP{Comparison to other approaches.} We have tried our best to experimentally 
%compare \sys to existing works in data race detection including 
%RELAY~\cite{Voung:FSE07}, WHOOP~\cite{Deligiannis:ASE15} and 
%Goblint~\cite{Vojdani:ASE16}. Unfortunately, they are too old and generate too 
%many errors when compiling and running, and thus we compare \sys to them 
%methodologically.

\PP{Limitations and future works.} \sys can be improved in some aspects. First, 
our alias-aware mining method only considers accessed variables and protecting 
locks that exist in the same data structure, and can miss some locking rules 
involving special locks such as RCU locks. To overcome this limitation, we plan 
to create a common virtual node for RCU lock in the alias graph, and regard it 
as the ancestor of all other nodes. In this way, all variables have a common 
ancestor with the RCU lock, and can be inferred to be protected by the RCU 
lock, because the accessed variable and the RCU lock are existing in the 
identical data structure in our alias-aware rule mining method. Second, \sys 
does not handle function-pointer calls, and thus it can not build complete call 
graphs, and thus can miss data races reached from function-pointer calls. To 
relieve this limitation, we plan to apply existing function-pointer 
analysis~\cite{Hind:TOPLAS99, Zhang:ISPEC15, Heintze:PLDI01} in \sys, to detect 
more data races in functions that are called through function pointers and 
reduce false negatives. Third, \sys performs a path-based analysis and suffers 
from high time overhead, to accelerate its analysis, we plan to employ 
summer-based analysis~\cite{Bai:ATC19, Bai:ATC22} to avoid re-analysis of the 
same function definition. Finally, we plan to apply \sys to detect data races 
in other OS kernels such as FreeBSD~\cite{FreeBSD} and NetBSD~\cite{NetBSD}, 
and use it to detect more types of concurrency bugs including use-after-free 
and uninitialized-variable access.

### 2023-08-02 7.related.tex
Finish a temporary version (without comparing experiment and some evaluation results)

Different from the above static analysis, \sys employs the alias graph to mine 
locking rules automatically. Benefiting from the path-based and field-sensitive 
alias relationships generated by alias graphs, \sys can deduce locking rules 
effectively. With these locking rules, \sys can detect data races accurately. 
Moreover, \sys exploits a pattern-based estimation to extract harmful data 
races automatically, and can help developers focus on data races that can 
indeed introduce serious issues.

@@

However, dynamic analysis suffers from low code coverage, and thus the locking 
rules inferred through logs can be imprecise. Besides, dynamic analyses need to 
run OS kernels with instrumentation and are hard to deploy.

>>>>

Different from the above static analysis, \sys employs the alias graph to mine 
locking rules automatically. Benefiting from the path-based and field-sensitive 
alias relationships generated by alias graphs, \sys can deduce locking rules 
effectively. With these locking rules, \sys can detect data races accurately. 
Moreover, \sys exploits a pattern-based estimation to extract harmful data 
races automatically, and can help developers focus on data races that can 
indeed introduce serious bugs.

@@

However, dynamic analysis suffers from low code coverage, and the locking rules 
inferred through execution traces such as logs can be imprecise. Besides, 
dynamic analyses need to run OS kernels with instrumentation and are hard to 
deploy.

### 2023-08-02 8.conclusion.tex
Finish a temporary version (without comparing experiment and some evaluation results)

>>>>

In this paper, we develop a novel static approach named \sys to detect data 
races in OS kernels by mining locking rules. It consists of three key 
techniques, including an alias-aware rule mining method to automatically deduce 
locking rules, a lock-usage analysis to filter out false positives caused by 
code that can not execute concurrently and a pattern-based estimation to 
extract harmful data races that can trigger memory or logic bugs such as 
null-pointer dereferences, data inconsistencies and double fetches. In the 
evaluation, \sys finds 88 real harmful data races, and 32 of them have been 
confirmed by the developers.

### 2023-08-02 paper.tex
Finish a temporary version (without comparing experiment and some evaluation results)

To take advantage of multiple-core architecture of modern CPUs, an operating 
system (OS) is expected to handle concurrency, inevitably introducing many 
concurrent issues. And data race is one of the most harmful concurrent issues 
that can trigger memory or logical bugs. A data race is caused by wrongly 
accessing shared data in different threads. To ensure correct access to shared 
data, fine-grained locking mechanisms are introduced. However, it is hard to 
determine whether a lock should be used for a specific variable (locking rules) 
even for an expert developer, due to poor documentation and complicated logic 
of OS code. As a result, OS kernel is prone to data races, because necessary 
locks may be missed by mistake. Static analysis is a common technique to help 
improve code quality, but it is quite challenging to detect data races in OS 
kernels automatically, because of lack of knowledge of locking rules and high 
complexity of concurrent execution.

In this paper, we design a practical static analysis approach named \sys, to 
effectively detect harmful races that can trigger memory or logical bugs in OS 
kernels by mining locking rules. \sys first employs an alias-aware rule mining 
method to automatically deduce locking rules, and detects data races caused by 
violation of these rules. And then performs a lock-usage analysis to filter out 
false positives caused by concurrency. At last, \sys extracts harmful data 
races from all detected data races through a pattern-based estimation. We have 
evaluated \sys on Linux 6.2, and find xxx data races, with a false positive 
rate of xxx. Among these data races, xxx are estimated to be harmful. We have 
reported these harmful bugs to Linux kernel developers, and xxx of them have 
been confirmed.

>>>>

To take advantage of multiple-core architecture of modern CPUs, an operating 
system (OS) is expected to handle concurrency, inevitably introducing many 
concurrency issues. And data race is one of the most harmful concurrency issues 
that can trigger memory or logic bugs. Data races are caused by wrongly 
accessing shared data in different threads. To ensure correct access to shared 
data, fine-grained locking mechanisms are introduced. However, it is hard to 
determine whether a lock should be held when accessing a specific variable 
(locking rules) even for an expert developer, because of poor document and 
complicated logic of OS code. As a result, OS kernel is prone to data races, 
because necessary locks may be missed by mistake. Static analysis is a common 
technique to improve code quality, but it is quite challenging to detect data 
races in OS kernels automatically, due to lack of knowledge of locking rules 
and high complexity of concurrent execution.

In this paper, we design a practical static analysis approach named \sys, to 
effectively detect data races and estimate their harmfulness in OS kernels by 
mining locking rules. \sys first employs an alias-aware rule mining method to 
automatically deduce locking rules, and detects data races caused by violation 
of these rules. And then performs a lock-usage analysis to filter out false 
positives caused by code that can not execute concurrently. At last, \sys 
extracts harmful data races from all detected data races through a 
pattern-based estimation. We have evaluated \sys on Linux 6.2, and find 273 
data races, with a false positive rate of 19.9\%. Among these data races, 88 
are estimated to be harmful. We have reported these harmful data races to Linux 
kernel developers, and 32 of them have been confirmed.
